Hot and fresh: Pervasive climate stressors of seagrass in a large Gulf coast estuary

### 2023-11-07 manu-draft.qmd
redo transect stressor metric models and figure

For comparison to seagrass, the annual metrics were also referenced to approximate periods of time between the annual seagrass transect surveys, as opposed to the calendar year as above. Transect surveys are typically done in the fall at the end of the growing season. The average date for each year across the subset of transects in each bay segment was estimated and the stressor metrics were referenced accordingly. For example, if the average transect date for a bay segment was September 15th in 2009 and Oct 1st in 2010, the 2010 stressor counts were based on all days between the two average dates. This was repeated for all years from 1998 to 2022 when transect data were available. Bay segment stressor metrics were calculated as the average counts in each "transect year" from all stations in each segment. The stressor metrics were compared to various response measures at the bay segments-scale from the transect data, including frequency occurrence each year (all species), percent change in frequency occurrence between years, and increase/decrease in frequency occurrence between years as a binary (1/0) variable. Generalized linear regression models were used to evaluate each of the seagrass response variables relative to the stressor metrics. Interaction factors were also included for bay segment and time period (recovery pre-2016, decline post-2016). The most parsimonious models were identified by forward and backward variable selection and comparison with AIC values [@sakamoto1986; @venables2002b]. All models excluded Lower Tampa Bay because of minimal seagrass change over time.

Additional models were constructed for the FIM and PDEM datasets to evaluate seagrass change relative to temperature and salinity. These models did not include stressor metrics, but rather used direct measurements of salinity and temperature as independent variables. Models of temperature or salinity to predict daily measurements were not possible with the FIM and PDEM monitoring designs because random locations were used for each sample date. However, the sample sizes for the FIM and PDEM models were much larger than those for the EPC data, where the latter was limited by the annual timestep of the transect surveys. This allowed for more flexible models that are less limited by degrees of freedom. GAMs were used to evaluate seagrass percent cover for the FIM data using continuous temperature and salinity as independent variables. Interactions with factor variables for bay segment and time period as above were also included. The GAMs also evaluated interactions between salinity and temperature to determine if both factors had an additive effect on seagrass change. The GAMs for the PDEM data were constructed similarly, except only Old Tampa Bay was evaluated (no bay segment interaction) and the seagrass response variable was presence/absence (i.e., a binomial distribution was used).

@@

![Ordination plots of Nonmetric Multidimensional Scaling results for annual estimates in each bay segment for total seagrass frequency occurrence (Freq. Occ.), counts of days above the temperature threshold (Temp.), counts of days below the salinity threshold (Sal.), and counts of days above or below for both, respectively. The left plot excludes counts of days with both temperature and salinity above and below the threshods, whereas the right plot excludes the separate temperature and salinity metrics. Points are sized by seagrass frequency occurrence. 95% confidence ellipses are shown for observations in each major bay segment. OTB: Old Tampa Bay, HB: Hillsborough Bay, MTB: Middle Tampa Bay.](figs/nms.png){#fig-nms}

![Predicted results from generalized linear models evaluating seagrass decline as (a) percent change between years and (b) probability of decline versus number of days when salinity was below the threshold. Models evaluated the response variable on the y-axes as a function of salinity days and interactions with time period (seagrass recovery prior to 2016, seagrass decline 2016 to present) and bay segment. Shaded areas are 95% confidence intervals. Colored lines in (b) as rug plots show instances of seagrass increase (blue) or decrease (red) by year. OTB: Old Tampa Bay, HB: Hillsborough Bay, MTB: Middle Tampa Bay.](figs/sgmod.png){#fig-sgmod}

>>>>

For comparison to seagrass, the annual metrics were also referenced to approximate periods of time between the annual seagrass transect surveys, as opposed to the calendar year as above. Transect surveys are typically done in the fall at the end of the growing season. The average date for each year across the subset of transects in each bay segment was estimated and the stressor metrics were referenced accordingly. For example, if the average transect date for a bay segment was September 15th in 2009 and Oct 1st in 2010, the 2010 stressor counts were based on all days between the two average dates. This was repeated for all years from 1998 to 2022 when transect data were available. Bay segment stressor metrics were calculated as the average counts in each "transect year" from all stations in each segment. The stressor metrics were compared to frequency occurrence (all species) each year by bay segment(all species). Generalized linear regression models were used to evaluate frequency occurrence in response to the independent variables, where the latter were the stressor metrics for temperature, salinity, both, bay segment, and time period (recovery pre-2016, decline post-2016). Two models wre evaluated, one with the temperature and salinity metrics and another with the both metrics given that the latter was a combined metric of the former. Interactions were included among all the predictor variables to test for changes in the relationships by bay segment, time period, and different levels of the stressor metrics. The most parsimonious models were identified by forward and backward variable selection and comparison with AIC values [@sakamoto1986; @venables2002b]. All models excluded Lower Tampa Bay because of minimal seagrass change over time.

Additional models were constructed for the FIM and PDEM datasets to evaluate seagrass change relative to temperature and salinity. These models used direct measurements of salinity and temperature as independent variables because the stressor metrics could not be calculated using the sampling designs from these monitoring programs. However, the sample sizes for the FIM and PDEM models were much larger than those for the EPC data, where the latter was limited by the annual timestep of the transect surveys. This allowed for more flexible models that are less limited by degrees of freedom. GAMs were used to evaluate seagrass percent cover for the FIM data using continuous temperature and salinity as independent variables. Interactions with factor variables for bay segment and time period as above were also included. The GAMs also evaluated interactions between salinity and temperature to determine if both factors had a combined effect on seagrass change. The GAMs for the PDEM data were constructed similarly, except only Old Tampa Bay was evaluated (no bay segment interaction) and the seagrass response variable was presence/absence (i.e., a binomial distribution was used).

@@

![Predicted results from generalized linear models evaluating seagrass frequency occurrence by time periods versus (a) the number of days when salinity was below the threshold and (b) the number of days when temperature was above the threshold. Time time periods were seagrass recovery prior to 2016 and seagrass decline 2016 to present. Shaded areas are 95% confidence intervals. Points are estimates of seagrass frequency and stressor metrics by bay segment. OTB: Old Tampa Bay, HB: Hillsborough Bay, MTB: Middle Tampa Bay.](figs/tempsalmod.png){#fig-sgmod}
